Exponential methods for solving hyperbolic problems with application to kinetic equations

### 2020-11-30 communication/20-12-04_canum/canum.tex
🖥  remove some images, and update presentation

\usepackage[backend=bibtex, style=authoryear-comp]{biblatex}
\usepackage{filecontents}
\newcommand{\customcite}[1]{\citeauthor{#1} (\citeyear{#1})}

@@

  Linear part is exact! \cmark
\end{frame}
%-------%
\begin{frame}{Idea of exponential integrators}
  \textbf{\color{mblue} 2 classes of methods:}
  \begin{description}
    \item[\textbf{exponential Runge-Kutta:}] solve exactly what we can, and interpolate the rest. For example first order exponential Euler method:
      $$
        u(t_n+\Delta t) \approx u^{n+1} = e^{-ia\Delta t}u^n + \Delta t\varphi_1(ia\Delta t)F(u^n)
      $$
      where $\varphi_1(z) = \dfrac{e^z - 1}{z}$
      \begin{thebibliography}{9}
        \setbeamertemplate{bibliography item}[article]
        \bibitem{a} \customcite{Hochbruck:2010}
      \end{thebibliography}
    \item[\textbf{Lawson:}] Change of variable: $v(t)=e^{-iat}u(t)$, we solve with a RK method: $\dot{v} = \tilde{F}(t,v) = e^{-iat}F(e^{iat}v(t))$

      For example, Lawson Euler method:
      $$
        v(t_n+\Delta t)\approx v^{n+1} = v^n + \Delta t e^{-iat_n}F(e^{iat_n}v^n)
      $$
      or as an expression of $u$:
      $$
        u^{n+1} = e^{-ia\Delta t}u^n + \Delta te^{ia\Delta t}F(u^n)
      $$
      \begin{thebibliography}{9}
        \setbeamertemplate{bibliography item}[article]
        \bibitem{a} \customcite{Isherwood:2018}
      \end{thebibliography}
  \end{description}
\end{frame}

@@

\subsection{Lawson methods}
%------------------------------------------------------------------------------
\begin{frame}{Lawson methods stability domain}
  For our toy model:
  $$
    \dot{u} = iau + \lambda(u)
  $$
  Change of variable: $v(t) = e^{-iat}u(t)$
  $$
    \dot{v} = e^{-iat}\lambda e^{iat}v
  $$
  Apply a Runge-Kutta method to compute stability function of Lawson method:
  $$
    v^{n+1} = \underbrace{p(\lambda\Delta t)}_{\text{stability function of RK}}v^n
  $$
  \emph{i.e.}:
  $$
    u^{n+1} = \overbrace{p(\lambda\Delta t)e^{-ia\Delta t}}^{\text{stability function of Lawson}}u^n
  $$
  Stability domain: $\mathcal{D}=\left\{z\in\mathbb{C},|p(z)|\leq 1\right\}$ of Lawson method is \textbf{the same} as the underlying Runge-Kutta method \textbf{because} $ia\in i\mathbb{R}$
\end{frame}
%-------%
\begin{frame}{Considered $Lawson(RK(s,p))$ methods}
  \begin{figure}\centering
    \includegraphics[height=0.8\textheight]{img/rk_sd.png}
  \end{figure}
\end{frame}
%-------%
\begin{frame}{Lawson methods -- CD2}
  For stability between a Lawson method and CD2, we solve:
  $$
    |p(iy)| = 1,\quad y\in\mathbb{R}
  $$
  \begin{table}
    \centering
    \begin{tabular}{|c|c|c|c|}
      \hline
      Methods & Lawson($RK(3,2) \; best$) & Lawson($RK(3,3)$) & Lawson($RK(4,4)$) \\
      \hline
      $y_{\max}$ & $2$ & $\sqrt{3}$ & $2\sqrt{2}$\\
      \hline  
    \end{tabular}
    \caption{CFL number for some Lawson schemes}
  \end{table}
\begin{thebibliography}{9}
  \setbeamertemplate{bibliography item}[article]
  \bibitem{} \customcite{Baldauf:2008}
\end{thebibliography}
\end{frame}
%-------%
\begin{frame}{Lawson methods -- LW5}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \includegraphics[width=\textwidth]{img/cfl_scheme}
    \end{column}
    \begin{column}{0.5\textwidth}
      \includegraphics[width=\textwidth]{img/cfl_rk44_weno.png}
    \end{column}
  \end{columns}
\end{frame}
%-------%
\begin{frame}{Lawson methods -- LW5: CFL estimates}
  \begin{table}
    \centering
    \begin{tabular}{|c|c|c|c|}
      \hline
      Methods & Lawson($RK(3,2) \; best$) & Lawson($RK(3,3)$) & Lawson($RK(4,4)$) \\
      \hline
      $\sigma $ & $1.344$ & $1.433$   & $1.73$   \\
      \hline  
    \end{tabular}
    \caption{CFL number for some Lawson schemes.}
  \end{table}
\begin{thebibliography}{9}
  \setbeamertemplate{bibliography item}[article]
  \bibitem{} \customcite{Motamed:2010}
  \bibitem{} \customcite{Lunet:2017}
\end{thebibliography}
\end{frame}

@@

\begin{frame}{Exponential Runge-Kutta methods}
  $$
    \dot{u} = iau + F(u)
  $$
  Example on ExpRK(2,2):
  $$
    \begin{aligned}
      u^{(1)} &= e^{-ia\Delta t}u^n - \Delta t\varphi_1 F(u^n) \\
      u^{n+1} &= e^{-ia\Delta t}u^n - \Delta t\left[ (\varphi_1-\varphi_2)F(u^n) + \varphi_2F(u^{(1)}) \right] 
    \end{aligned}
  $$
  % $$
  %   \begin{aligned}
  %     u^{(1)} &= u^n(1+ia\varphi_{1,2}) + ia\Delta t\varphi_{1,2}F(u^n) \\
  %     u^{n+1} &= u^n(1+ia\Delta t\varphi_1) + \frac{\Delta t}{2}\varphi_1F(u^n)+\frac{\Delta t}{2}\varphi_1F(u^{(1)})
  %   \end{aligned}
  % $$

@@

  \mbold{Numerical tools:}
  \begin{itemize}
    \item FFT in $x$ direction
    \item CD2 or WENO5 in $v$ direction
    \item $Lawson(RK(s,p))$ or ExpRK method in time $t$
  \end{itemize}

@@

  Because of damping:
  $$
    \max_n||E^n||_{\infty} = ||E^0||_\infty
  $$
  So, we choose $\Delta t = 0.1$ (with $\Delta t = 100$ it is still stable!)
\end{frame}
%-------%
\begin{frame}{Landau damping: numerical results}
  \only<1>{
    \begin{figure}
      \centering
      \resizebox{!}{.7\paperheight}{\input{img/hn}}
      \caption{Landau damping test: time history of the CFL condition (semi-log scale).}
      \label{ld}
    \end{figure}
  }\only<2>{
    \begin{figure}
      \centering
      \resizebox{!}{.7\paperheight}{\input{img/Emax}}
      \caption{Landau damping test: time history of $\|E(t)\|_{L^2}$ (semi-log scale) obtained with Lawson($RK(4, 4)$) and WENO5 
      with $\Delta t=1/8$ and $\Delta t=1$.}
      \label{ld}
    \end{figure}
  }
\end{frame}
%-------%
\begin{frame}{Bump on Tail (BoT)}
  $$
    f(t=0,x,v) = \left[\frac{0.9}{\sqrt{2\pi}}e^{-\frac{v^2}{2}} + \frac{0.2}{\sqrt{2\pi}}e^{-2(v-4.5)^2} \right](1+0.001\cos(0.5x))
  $$

@@

  Numerical estimation of $\max_n||E^n||_\infty\approx 0.6$, we choose $\Delta t = \frac{C\Delta v}{0.6}$
\end{frame}
%-------%
\begin{frame}{BoT: numerical results}
  \only<1>{
    \begin{figure}
    \centering
        \includegraphics[width=\textwidth]{img/vp_cfl.png}
        \caption{Distribution function at time $t=40$ as a function of $x$ and $v$ for Lawson($RK(4, 4)$) + WENO5 (left), Lawson($RK(4, 4)$) + centered scheme (center), Hochbruck--Ostermann + centered scheme (right).}  
    \label{space}      
    \end{figure}
  } \only<2> {
    \begin{columns}
      \begin{column}{0.5\textwidth}
        \begin{figure}
          \includegraphics[width=0.9\textwidth]{img/ee_weno_rk44.png}
          \caption{Illustration of the accuracy of the CFL estimate obtained from the linear theory. History of electric energy with Lawson($RK(4,4)$) + WENO5}
        \end{figure}
        \vfill
        \ 
      \end{column}
      \begin{column}{0.5\textwidth}
        \begin{figure}
          \includegraphics[width=0.9\textwidth]{img/bot_cfl_weno_rk44.png}
          \caption{History of CFL condition for Lawson($RK(4,4)$) + WENO5 case}
        \end{figure}
        \vfill
        \ 
      \end{column}
    \end{columns}
  } \only<3> {
    \begin{figure}
      \includegraphics[width=0.9\textwidth]{img/H.png}
    \end{figure}
  }
\end{frame}

@@

\begin{frame}{Conclusion}
  \mbold{Summary}
  \begin{itemize}
    \item Better understanding on stability of Lawson or ExpRK methods in transport equations
    \item Python script with \texttt{sympy} to compute estimates of CFL of Lawson -- CD2, Lawson -- WENO (5 or 3) or ExpRK -- CD2 (with relaxing CFL)
    \item An adaptive time step size which works with any time integrators
  \end{itemize}

  \mbold{Future works}
  \begin{itemize}
    \item We can improve method with an embedded Runge-Kutta method (Dormand-Prince method, used in \texttt{ode45} of Matlab)
    \item Compare performance between exponential integrators and splitting methods (same stages/step, same order?)
    \item Use semi-Lagrangian method to remove dependency on periodic space (Fourier transform)
  \end{itemize}
\end{frame}

>>>>

\usepackage[backend=bibtex, style=authoryear-comp]{biblatex}
\usepackage{filecontents}
\renewbibmacro*{cite}{%
  \iffieldundef{shorthand}
    {\ifthenelse{\ifnameundef{labelname}\OR\iffieldundef{labelyear}}
       {\usebibmacro{cite:label}%
        \setunit{\printdelim{nonameyeardelim}}}
       {\printnames{labelname}%
        \setunit{\printdelim{nameyeardelim}}}%
     \usebibmacro{cite:labeldate+extradate}%
     \setunit{\addcomma\space}%
     \usebibmacro{journal}}
    {\usebibmacro{cite:shorthand}}}
\newcommand{\customcite}[1]{\cite{#1}}

@@

  Linear part is exact! \cmark
\end{frame}
%-------%
\begin{frame}{Idea of exponential integrators}
  \textbf{\color{mblue} 2 classes of methods:}
  \begin{description}
    \item[\textbf{exponential Runge-Kutta:}] solve exactly what we can, and interpolate the rest. For example first order exponential Euler method:
      $$
        u(t_n+\Delta t) \approx u^{n+1} = e^{-ia\Delta t}u^n + \Delta t\varphi_1(ia\Delta t)F(u^n)
      $$
      where $\varphi_1(z) = \dfrac{e^z - 1}{z}$
      \vspace{-0.1cm}
      \begin{thebibliography}{9}
        \setbeamertemplate{bibliography item}[article]
        \bibitem{a} \customcite{Hochbruck:2010}
      \end{thebibliography}
    \item[\textbf{Lawson:}] Change of variable: $v(t)=e^{-iat}u(t)$, we solve with a RK method: $\dot{v} = \tilde{F}(t,v) = e^{-iat}F(e^{iat}v(t))$

      For example, Lawson Euler method:
      $$
        v(t_n+\Delta t)\approx v^{n+1} = v^n + \Delta t e^{-iat_n}F(e^{iat_n}v^n)
      $$
      or as an expression of $u$:
      $$
        u^{n+1} = e^{-ia\Delta t}u^n + \Delta te^{ia\Delta t}F(u^n)
      $$
      \vspace{-0.75cm}
      \begin{thebibliography}{9}
        \setbeamertemplate{bibliography item}[article]
        \bibitem{a} \customcite{Isherwood:2018}
        %\bibitem{a} \cite{Isherwood:2018}
      \end{thebibliography}
  \end{description}
\end{frame}

@@

\subsection{Lawson methods}
%------------------------------------------------------------------------------
\begin{frame}{Lawson methods stability domain}
  For our toy model:
  $$
    \dot{u} = iau + \lambda(u)
  $$
  Change of variable: $v(t) = e^{-iat}u(t)$
  $$
    \dot{v} = e^{-iat}\lambda e^{iat}v
  $$
  Apply a Runge-Kutta method to compute stability function of Lawson method:
  $$
    v^{n+1} = \underbrace{p(\lambda\Delta t)}_{\text{stability function of RK}}v^n
  $$
  \emph{i.e.}:
  $$
    u^{n+1} = \overbrace{p(\lambda\Delta t)e^{-ia\Delta t}}^{\text{stability function of Lawson}}u^n
  $$
  Stability domain: $\mathcal{D}=\left\{z\in\mathbb{C},|p(z)|\leq 1\right\}$ of Lawson method is \textbf{the same} as the underlying Runge-Kutta method \textbf{because} $ia\in i\mathbb{R}$
\end{frame}
%-------%
\begin{frame}{Considered $Lawson(RK(s,p))$ methods}
  \begin{figure}\centering
    \includegraphics[height=0.8\textheight]{img/rk_sd.png}
  \end{figure}
\end{frame}
%-------%
\begin{frame}{Lawson methods -- CD2}
  For stability between a Lawson method and CD2, we solve:
  $$
    |p(iy)| = 1,\quad y\in\mathbb{R}
  $$
\end{frame}
%-------%
\begin{frame}{Lawson methods -- LW5}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \includegraphics[width=\textwidth]{img/cfl_scheme}
    \end{column}
    \begin{column}{0.5\textwidth}
      \includegraphics[width=\textwidth]{img/cfl_rk44_weno.png}
    \end{column}
  \end{columns}
\end{frame}
%-------%
\begin{frame}{Lawson methods -- CD2/LW5: CFL estimates}
  \begin{table}
    \centering
    \begin{tabular}{|c|c|c|c|}
      \hline
      - & Lawson($RK(3,3)$) & Lawson($RK(4,4)$) \\
      \hline
      CD2 ($y_{\max}$) & $\sqrt{3}$ & $2\sqrt{2}$\\
      \hline
      LW5 ($\sigma$) & $1.433$   & $1.73$   \\
      \hline  
    \end{tabular}
    \caption{CFL number for some Lawson schemes.}
  \end{table}
  Other RK--CD2 CFL estimates:
\begin{thebibliography}{9}
  \setbeamertemplate{bibliography item}[article]
  \bibitem{} \customcite{Baldauf:2008}
\end{thebibliography}
  Same estimation of RK--WENO CFL estimates:
\begin{thebibliography}{9}
  \setbeamertemplate{bibliography item}[article]
  \bibitem{} \customcite{Motamed:2010}
  \bibitem{} \customcite{Lunet:2017}
\end{thebibliography}
\end{frame}

@@

\begin{frame}{Exponential Runge-Kutta methods}
  $$
    \dot{u} = iau + F(u)
  $$
  Example on ExpRK(2,2):
  $$
    \begin{aligned}
      u^{(1)} &= e^{-ia\Delta t}u^n - \Delta t\varphi_1 F(u^n) \\
      u^{n+1} &= e^{-ia\Delta t}u^n - \Delta t\left[ (\varphi_1-\varphi_2)F(u^n) + \varphi_2F(u^{(1)}) \right] 
    \end{aligned}
  $$

@@

  \mbold{Numerical tools:}
  \begin{itemize}
    \item FFT in $x$ direction
    \item WENO5 in $v$ direction
    \item $Lawson(RK(s,p))$ method in time $t$
  \end{itemize}

@@

  Because of damping:
  $$
    \max_n||E^n||_{\infty} = ||E^0||_\infty
  $$
  So, we choose $\Delta t = 0.1$ (with $\Delta t = 100$ it is still stable!)
\end{frame}
%-------%
\begin{frame}{Landau damping: numerical results}
  \only<1>{
    \begin{figure}
      \centering
      \resizebox{!}{.7\paperheight}{\input{img/Emax}}
      \caption{Landau damping test: time history of $\|E(t)\|_{L^2}$ (semi-log scale) obtained with Lawson($RK(4, 4)$) and WENO5 
      with $\Delta t=1/8$ and $\Delta t=1$.}
      \label{ld}
    \end{figure}
  }
  \only<2>{
    \begin{figure}
      \centering
      \resizebox{!}{.7\paperheight}{\input{img/hn}}
      \caption{Landau damping test: time history of the CFL condition (semi-log scale).}
      \label{ld}
    \end{figure}
  }
\end{frame}
%-------%
\begin{frame}{Bump on Tail (BoT)}
  $$
    f(t=0,x,v) = \left[\frac{0.9}{\sqrt{2\pi}}e^{-\frac{v^2}{2}} + \frac{0.2}{\sqrt{2\pi}}e^{-2(v-4.5)^2} \right](1+0.001\cos(0.5x))
  $$

@@

  Numerical estimation of $\max_n||E^n||_\infty\approx 0.6$, we choose $\Delta t = \frac{C\Delta v}{0.6}$
\end{frame}
%-------%
\begin{frame}{BoT: numerical results}
  \only<1>{
    \begin{figure}
    \centering
        \includegraphics[width=\textwidth]{img/vp_cfl.png}
        \caption{Distribution function at time $t=40$ as a function of $x$ and $v$ for Lawson($RK(4, 4)$) + WENO5 (left), Lawson($RK(4, 4)$) + centered scheme (center), Hochbruck--Ostermann + centered scheme (right).}  
    \label{space}      
    \end{figure}
  } \only<2> {
    \begin{columns}
      \begin{column}{0.5\textwidth}
        \begin{figure}
          \includegraphics[width=0.9\textwidth]{img/ee_weno_rk44.png}
          \caption{Illustration of the accuracy of the CFL estimate obtained from the linear theory. History of electric energy with Lawson($RK(4,4)$) + WENO5}
        \end{figure}
        \vfill
        \ 
      \end{column}
      \begin{column}{0.5\textwidth}
        \begin{figure}
          \includegraphics[width=0.9\textwidth]{img/bot_cfl_weno_rk44.png}
          \caption{History of CFL condition for Lawson($RK(4,4)$) + WENO5 case}
        \end{figure}
        \vfill
        \ 
      \end{column}
    \end{columns}
  }
\end{frame}

@@

\begin{frame}{Conclusion}
  \mbold{Summary}
  \begin{itemize}
    \item Better understanding on stability of Lawson or ExpRK methods in transport equations
    \item Python script with \texttt{sympy} to compute estimates of CFL of Lawson -- CD2, Lawson -- WENO (5 or 3)
    %\item An adaptive time step size which works with any time integrators
  \end{itemize}

  \mbold{Future works}
  \begin{itemize}
    \item We can improve method with an embedded Runge-Kutta method (Dormand-Prince method, used in \texttt{ode45} of Matlab)
    \item Compare performance between exponential integrators and splitting methods (same order)
    \item Use semi-Lagrangian method to remove dependency on periodic space (Fourier transform)
  \end{itemize}
\end{frame}
